Brother, Can You Spare a Manufacturing Job?  How Voters React to Deindustrialization

### 2021-11-30 rCode/Draft.Rmd
tweaks

---
title: "Brother, can you spare a manufacturing job? White Voters reaction to Deindustrialization"
author: 'Catherine Darin and Zagreb Mukerjee'
date: '11/29/2021'
fontsize: 12pt
output:
  pdf_document: default
urlcolor: red
header-includes:
  - \usepackage{bm}
---

@@

What led to Donald Trump's surprising 2016 election victory? This paper examines the potential contribution of deindustrialization. Counties differ by exposure to manufacturing, which we use in an instrumental-variable approach to identify the effect of manufacturing job loss on change in Democratic vote share. We find that Democratic vote share falls in counties with high manufacturing job loss since 2004. Disaggregating this result by race shows that the democratic vote share fell where layoffs affected white populations, and rose where layoffs affected nonwhites. This suggests a racial component to how voters process economic hardship.

@@

To do: Qualitative description of the trajectory of XYZ county that swung from O to T and suffered big manufacturing layoffs. This will establish the stakes and connect this to the broader conversation.

@@

Why did Trump win in 2016? A longstanding debate about relative importance of race and economic factors. Complicated by the interaction of the two. We hope to identify the importance of manufacturing job losses, both in aggregate and for different racial groups.

@@

In their main model, Baccini and Weymouth looked at the effect on gross manufacturing job losses from 2012 to 2015 on the change in Democratic vote share in the 2016 presidential election.

Our key extension of Baccini and Weymouth's analysis is changing the quantity of interest from **gross** manufacturing job losses to **net** manufacturing job losses. For example, in Baccini and Weymouth's model, if a county lost 400 manufacturing jobs from 2012-2015 and also gained 400 manufacturing jobs over the same period, they would count 400 overall layoffs. Using our measure of *net* manufacturing job losses, we would consider this example a 0 net change in manufacturing employment. Given that job destruction and creation is a natural dynamic within the economy (i.e., creative destruction), we think that *net* job losses is a more meaningful measure.

Using *net* manufacturing job losses as our outcome, we replicate Baccini and Weymouth's main model, which uses 2012-2015 manufacturing (gross) job losses to explain the change in democratic vote share in 2016. Using their specifications, we find totally opposite results - that a *net* increase in white manufacturing jobs *increased* democratic vote share, while a *net* increase in non-white manufacturing jobs *decreased* democratic vote share. (See Table 3)

However, we do not trust the result of this specification, because, from 2012 to 2015, the national economy experienced a net increase in manufacturing jobs following the 2008 recession (See Figure 1). During this time of recovery, counties that experienced the most pronounced manufacturing job losses in prior years were also likely to experience the most pronounced recovery. Therefore, because manufacturing job losses pre-2012 are likely negatively correlated with manufacturing job gains from 2012-2015, it is misleading to segment the analysis data to just include job losses from 2012 to 2015. [To DO: Empirically show negative correlation describe here]

If we expand our measure of net manufacturing job losses to include the 2004 to 2011 period - which covers most of the period when the U.S. manufacturing economy was hit most intensely (See figure 1) -- we get a result that is consistent with Baccini and Weymouth's overall conclusion (See Table 4). That is, we find that net white manufacturing job losses decreased democratic vote share, while net non-white manufacturing job losses increased democratic vote share.

While our main takeaway is consistent with Baccini and Weymouth's story, it is important to highlight the substantive importance of the *timing* of manufacturing layoffs on political outcomes. Whereas Baccini and Weymouth's analysis suggests that economic changes lead to immediate political consequences, our analysis shows that economic changes can have lagged and variable effects on political outcomes.

@@

We find that, in aggregate, manufacturing job losses were a driver of change in Democratic vote share in the 2016 election. Breaking it down by race, we see that

Moreover, we find that, given the evolution of the business cycle, considering only recent job losses can paint a misleading picture of the relationship between manufacturing job losses and changes in vote share. In fact, we find that net manufacturing job losses from 2004 to 2011 are crucial to explaining Trump's 2016 win. This finding suggests that economic changes, like manufacturing layoffs, can have lagged and variable political consequence.

@@

### Manufacturing Layoffs Overtime

```{r}
# manufacturing layoffs overtime

@@

national_ts <- national_ts %>%
  rename(year = 1,
         net_job_change = 2) %>%
  dplyr::select(year, net_job_change) %>%
  mutate(
    year = as.numeric(year),
    net_job_change = net_job_change*4) %>% group_by()

ggplot(data = national_ts %>% filter(year >= 1995), aes(x = year, y = net_job_change)) +
  geom_col() + scale_x_continuous(breaks=seq(1995,2019,2)) +
	scale_y_continuous(labels = label_number(suffix = "", scale = 1e-6)) +
  labs()

@@

### Example

@@

rownames(regResultList$secondStageModelNetCA$coefficients )[rownames(regResultList$secondStageModelNetCA$coefficients)=="firstStageModelNetCA$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelNetC$coefficients )[rownames(regResultList$secondStageModelNetC$coefficients)=="firstStageModelNetC$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetCA$coefficients )[rownames(regResultList$secondStageModelWNetCA$coefficients)=="firstStageModelWNetCA$fitted.values"] <-  "White Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC$coefficients )[rownames(regResultList$secondStageModelWNetC$coefficients)=="firstStageModelWNetC$fitted.values"] <-  "White Manufacturing Layoffs"

@@

rownames(regResultList$secondStageModelNetCA$beta )[rownames(regResultList$secondStageModelNetCA$beta) =="firstStageModelNetCA$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelNetC$beta )[rownames(regResultList$secondStageModelNetC$beta)=="firstStageModelNetC$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetCA$beta )[rownames(regResultList$secondStageModelWNetCA$beta)=="firstStageModelWNetCA$fitted.values"] <-  "White Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC$beta )[rownames(regResultList$secondStageModelWNetC$beta)=="firstStageModelWNetC$fitted.values"] <-  "White Manufacturing Layoffs"

@@

rownames(regResultList$secondStageModelNetC04A$coefficients )[rownames(regResultList$secondStageModelNetC04A$coefficients)=="firstStageModelNetC04A$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelNetC04$coefficients )[rownames(regResultList$secondStageModelNetC04$coefficients)=="firstStageModelNetC04$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC04A$coefficients )[rownames(regResultList$secondStageModelWNetC04A$coefficients)=="firstStageModelWNetC04A$fitted.values"] <-  "White Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC04$coefficients )[rownames(regResultList$secondStageModelWNetC04$coefficients)=="firstStageModelWNetC04$fitted.values"] <-  "White Manufacturing Layoffs"

rownames(regResultList$secondStageModelNetC04A$beta )[rownames(regResultList$secondStageModelNetC04A$beta) =="firstStageModelNetC04A$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelNetC04$beta )[rownames(regResultList$secondStageModelNetC04$beta)=="firstStageModelNetC04$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC04A$beta )[rownames(regResultList$secondStageModelWNetC04A$beta)=="firstStageModelWNetC04A$fitted.values"] <-  "White Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC04$beta )[rownames(regResultList$secondStageModelWNetC04$beta)=="firstStageModelWNetC04$fitted.values"] <-  "White Manufacturing Layoffs"

>>>>

---
title: "Brother, Can You Spare a Manufacturing Job?  How Voters React to Deindustrialization"
author: 'Catherine Darin and Zagreb Mukerjee'
date: '11/29/2021'
fontsize: 12pt
output:
  pdf_document: default
urlcolor: red
header-includes:
  - \usepackage{bm}
---

@@

What led to Donald Trump's surprising 2016 election victory? This paper examines the potential contribution of deindustrialization. Counties differ by exposure to manufacturing, which we use in an instrumental-variable approach to identify the effect of manufacturing job loss on change in Democratic vote share. We find that Democratic vote share falls in counties with high manufacturing job loss since 2004. Disaggregating this result by race shows that the Democratic vote share fell where layoffs affected white populations, and rose where layoffs affected nonwhites. This suggests a racial component to how voters process economic hardship.

@@

To do: Qualitative description of the trajectory of XYZ county that swung from Obama to Trump and suffered big manufacturing layoffs. This will establish the stakes and connect this to the broader conversation.

@@

Why did Trump win in 2016? There is a longstanding debate about relative importance of race and economic factors. Complicated by the interaction of the two. We hope to identify the importance of manufacturing job losses, both in aggregate and for different racial groups. Eventually we think that people should stop asking "race or economics" when the answer is likely both.

@@

In their main model, Baccini and Weymouth looked at the effect of gross manufacturing job losses from 2012 to 2015 on the change in Democratic vote share in the 2016 presidential election.

Our key extension of Baccini and Weymouth's analysis is changing the quantity of interest from **gross** manufacturing job losses to **net** manufacturing job losses. For example, in Baccini and Weymouth's model, if a county lost 400 manufacturing jobs from 2012-2015 and also gained 400 manufacturing jobs over the same period, they would count 400 overall layoffs. Using our measure of *net* manufacturing job losses, we would consider this example a 0 net change in manufacturing employment. Given that job destruction and creation is a natural dynamic within the economy (i.e. seasonal employment, creative destruction), we think that *net* job losses is a more meaningful measure.

Using *net* manufacturing job losses as our outcome, we replicate Baccini and Weymouth's main model, which uses 2012-2015 manufacturing (gross) job losses to explain the change in Democratic vote share in 2016. Using their specifications, we find totally opposite results - that a *net* increase in white manufacturing jobs *increased* Democratic vote share, while a *net* increase in non-white manufacturing jobs *decreased* Democratic vote share. (See Table 3)

However, we don't think this is capturing the important dynamic at play. From 2012 to 2015, the national economy experienced a net increase in manufacturing jobs following the 2008 recession (See Figure 1). During this time of recovery, counties that experienced the most pronounced manufacturing job losses in prior years were also likely to experience the most pronounced recovery. Because of this v-shaped recovery, focusing only on 2012-2015 job losses/gains paints a distorted picture. [To DO: Empirically show negative correlation described here]

If we expand our measure of net manufacturing job losses to include the 2004 to 2011 period - which covers most of the period when the U.S. manufacturing economy was hit most intensely (See Figure 1) -- we get a result that corroborates the core intuition of Baccini and Weymouth (See Table 4). That is, we find that net white manufacturing job losses decreased Democratic vote share, while net non-white manufacturing job losses increased democratic vote share.

It is important to highlight the substantive importance of the *timing* of manufacturing layoffs on political outcomes. Whereas Baccini and Weymouth's analysis suggests that economic changes lead to immediate political consequences, our analysis shows that economic changes can have lagged and variable effects on political outcomes. The mechanisms of these lagged effects merit further study.

@@

We find that, in aggregate, manufacturing job losses were associated with loss of Democratic vote share in the 2016 election. 10 manufacturing job losses per worker in a county since 2004 corresponded to a 3.8 \% loss in vote share, plus or minus 3.3 \%. Breaking it down by race, we see that white

@@

### Nationwide Manufacturing Layoffs Per Worker, 1995-2019

```{r, message = F}
# manufacturing layoffs over time

@@

national_ts <- national_ts %>%
  rename(year = 1,
         net_job_change = 2) %>%
  dplyr::select(year, net_job_change) %>%
  rename(Year = year, `Job Change per Worker` = net_job_change) %>% 
  mutate(
    Year = as.numeric(Year),
    `Job Change per Worker` = `Job Change per Worker`*4) %>% group_by()

ggplot(data = national_ts %>% filter(Year >= 1995), aes(x = Year, y = `Job Change per Worker`)) +
  geom_col() + scale_x_continuous(breaks=seq(1995,2019,2)) +
	scale_y_continuous(labels = label_number(suffix = "", scale = 1e-6)) +
  labs()

@@

@@

rownames(regResultList$secondStageModelNetC04A$coefficients )[rownames(regResultList$secondStageModelNetC04A$coefficients)=="firstStageModelOurs04_2$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelNetC04$coefficients )[rownames(regResultList$secondStageModelNetC04$coefficients)=="firstStageModelOurs04_3$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC04A$coefficients )[rownames(regResultList$secondStageModelWNetC04A$coefficients)=="firstStageModelOurs04_5$fitted.values"] <-  "White Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC04$coefficients )[rownames(regResultList$secondStageModelWNetC04$coefficients)=="firstStageModelOurs04_6$fitted.values"] <-  "White Manufacturing Layoffs"

rownames(regResultList$secondStageModelNetC04A$beta )[rownames(regResultList$secondStageModelNetC04A$beta) =="firstStageModelOurs04_2$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelNetC04$beta )[rownames(regResultList$secondStageModelNetC04$beta)=="firstStageModelOurs04_3$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC04A$beta )[rownames(regResultList$secondStageModelWNetC04A$beta)=="firstStageModelOurs04_5$fitted.values"] <-  "White Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC04$beta )[rownames(regResultList$secondStageModelWNetC04$beta)=="firstStageModelOurs04_6$fitted.values"] <-  "White Manufacturing Layoffs"

rownames(regResultList$secondStageModelNetCA$coefficients )[rownames(regResultList$secondStageModelNetCA$coefficients)=="firstStageModelOurs_2$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelNetC$coefficients )[rownames(regResultList$secondStageModelNetC$coefficients)=="firstStageModelOurs_3$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetCA$coefficients )[rownames(regResultList$secondStageModelWNetCA$coefficients)=="firstStageModelOurs_5$fitted.values"] <-  "White Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC$coefficients )[rownames(regResultList$secondStageModelWNetC$coefficients)=="firstStageModelOurs_6$fitted.values"] <-  "White Manufacturing Layoffs"

@@

rownames(regResultList$secondStageModelNetCA$beta )[rownames(regResultList$secondStageModelNetCA$beta) =="firstStageModelOurs_2$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelNetC$beta )[rownames(regResultList$secondStageModelNetC$beta)=="firstStageModelOurs_3$fitted.values"] <-  "Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetCA$beta )[rownames(regResultList$secondStageModelWNetCA$beta)=="firstStageModelOurs_5$fitted.values"] <-  "White Manufacturing Layoffs"
rownames(regResultList$secondStageModelWNetC$beta )[rownames(regResultList$secondStageModelWNetC$beta)=="firstStageModelOurs_6$fitted.values"] <-  "White Manufacturing Layoffs"

@@
